Constraints on Cosmic-Ray Particle Populations, Magnetic Fields, and Dark Matter from VHE Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS

### 2011-09-01 ms.tex
Citations updated

Moreover, a substantial value of $X_\CR$ can bias hydrostatic cluster masses as well as the temperature decrement of the CMB in the direction of a galaxy cluster due to the Sunyaev-Zel'dovich effect which is proportional to the line-of-sight integral of the thermal electron pressure. This could then severely jeopardize the use of clusters to determine cosmological parameters. Comparing X-ray and optical potential profiles in the centers of galaxy clusters yields an upper limit of 20--30 \% of non-thermal pressure (that can be composed of CRs, magnetic fields or turbulence) relative to the thermal gas pressure \citep{2008MNRAS.388.1062C, 2010MNRAS.404.1165C}. An analysis that compares spatially resolved weak gravitational lensing and hydrostatic X-ray masses for a sample of 18 galaxy clusters detects a deficit of the hydrostatic in comparison to the lensing mass of $20 \%$ at $R_{500}$ -- within which the
mean density is 500 times the critical density -- suggesting again a substantial non-thermal pressure contribution on large scales \citep{2008MNRAS.384.1567M}. High-energy gamma-ray emission is a complementary method of constraining the pressure contribution of CRs that is most sensitive
to the cluster core region. However, it assumes that the CR component is fully mixed with the ICM and almost insensitive in detecting a two-phase structure of CRs and the thermal ICM.  A fraction $X_\CR$ of only a few percent is required in order to produce a gamma-ray flux observable with current generation
gamma-ray telescopes rendering this technique at least as sensitive as the dynamical and hydrostatic method (which are however more general in that they are sensitive to any non-thermal pressure component).

@@

A third mechanism for gamma-ray production in a galaxy cluster is self-annihilation of a DM particle, e.g. a weakly interacting massive particle (WIMP). As already mentioned, about 80\% of the cluster mass is in the form of dark matter. Recent works \citep{article:EvansFerrerSarkar:2004, article:BergstromHooper:2006} have shown that the gamma-ray emissivity from DM annihilations is expected to be proportional to the square of the DM density integrated along the line of sight, making clusters of galaxies potentially good
candidates for DM searches \citep{article:PinzkePfrommerBergstrom2009, 2011ApJ...726L...6C}. This despite their large distances compared to other common targets for DM searches, such as dwarf spheroidal galaxies \citep{article:Strigari_etal:2007, article:Acciari_etal:2010,article:Aliu_etal:2009} or the Galactic center \citep{article:Kosack_etal:2004, article:Aharonian_etal:2006, article:Aharonian_etal:2009b}.

@@

The Coma cluster of galaxies (ACO 1656) is one of the most thoroughly studied clusters over all wavelengths \citep{article:Voges_etal:1999}. Located at a distance of about 100 Mpc \citep[$z=0.023$;][]{article:StrubleRood:1999}, it is one of the closest massive clusters \citep[$M \sim10^{15}M_{\odot}$;][]{article:Smith:1983, article:Kubo_etal:2008}. It hosts both a giant radio halo
\citep{article:Giovannini_etal:1993,article:Thierbach_etal:2003} and peripheral radio relic, which appears connected to the radio halo with a ``diffuse'' bridge \citep[see discussion in][]{article:BrownRudnick:2010}. It has been suggested that the relic is an infall shock \citep{article:Ensslin_etal:1998} and successively demonstrated by cosmological simulations that model the non-thermal emission processes \citep{article:PfrommerEnsslinSpringel:2008, article:Pfrommer:2008, article:Battaglia_etal:2009, article:Skillman_etal:2011}. Extended soft thermal X-ray (SXR) emission is evident from the ROSAT all-sky survey in the 0.1 to 2.4 keV band \citep{article:BrielHenryBohringer:1992} and observations with XMM-Newton \citep{article:Briel_etal:2001} revealed substructure in the X-ray halo with substantial turbulent pressure support that shows a lower limit of $\sim 10 \%$ of the total intracluster medium pressure \citep{2004A&A...426..387S}. As such it is a natural candidate for gamma-ray observations.

>>>>

Moreover, a substantial value of $X_\CR$ can bias hydrostatic cluster masses as well as the temperature decrement of the CMB in the direction of a galaxy cluster due to the Sunyaev-Zel'dovich effect which is proportional to the line-of-sight integral of the thermal electron pressure. This could then severely jeopardize the use of clusters to determine cosmological parameters. Comparing X-ray and optical potential profiles in the centers of galaxy clusters yields an upper limit of 20--30 \% of non-thermal pressure (that can be composed of CRs, magnetic fields or turbulence) relative to the thermal gas pressure \citep{article:Churazov_etal:2008, article:Churazov_etal:2010}. An analysis that compares spatially resolved weak gravitational lensing and hydrostatic X-ray masses for a sample of 18 galaxy clusters detects a deficit of the hydrostatic in comparison to the lensing mass of $20 \%$ at $R_{500}$ -- within which the
mean density is 500 times the critical density -- suggesting again a substantial non-thermal pressure contribution on large scales \citep{article:Mahdavi_etal:2008}. High-energy gamma-ray emission is a complementary method of constraining the pressure contribution of CRs that is most sensitive
to the cluster core region. However, it assumes that the CR component is fully mixed with the ICM and almost insensitive in detecting a two-phase structure of CRs and the thermal ICM.  A fraction $X_\CR$ of only a few percent is required in order to produce a gamma-ray flux observable with current generation
gamma-ray telescopes rendering this technique at least as sensitive as the dynamical and hydrostatic method (which are however more general in that they are sensitive to any non-thermal pressure component).

@@

A third mechanism for gamma-ray production in a galaxy cluster is self-annihilation of a DM particle, e.g. a weakly interacting massive particle (WIMP). As already mentioned, about 80\% of the cluster mass is in the form of dark matter. Recent works \citep{article:EvansFerrerSarkar:2004, article:BergstromHooper:2006} have shown that the gamma-ray emissivity from DM annihilations is expected to be proportional to the square of the DM density integrated along the line of sight, making clusters of galaxies potentially good
candidates for DM searches \citep{article:PinzkePfrommerBergstrom2009, article:Cuesta_etal:2011}. This despite their large distances compared to other common targets for DM searches, such as dwarf spheroidal galaxies \citep{article:Strigari_etal:2007, article:Acciari_etal:2010,article:Aliu_etal:2009} or the Galactic center \citep{article:Kosack_etal:2004, article:Aharonian_etal:2006, article:Aharonian_etal:2009b}.

@@

The Coma cluster of galaxies (ACO 1656) is one of the most thoroughly studied clusters over all wavelengths \citep{article:Voges_etal:1999}. Located at a distance of about 100 Mpc \citep[$z=0.023$;][]{article:StrubleRood:1999}, it is one of the closest massive clusters \citep[$M \sim10^{15}M_{\odot}$;][]{article:Smith:1983, article:Kubo_etal:2008}. It hosts both a giant radio halo
\citep{article:Giovannini_etal:1993,article:Thierbach_etal:2003} and peripheral radio relic, which appears connected to the radio halo with a ``diffuse'' bridge \citep[see discussion in][]{article:BrownRudnick:2010}. It has been suggested that the relic is an infall shock \citep{article:Ensslin_etal:1998} and successively demonstrated by cosmological simulations that model the non-thermal emission processes \citep{article:PfrommerEnsslinSpringel:2008, article:Pfrommer:2008, article:Battaglia_etal:2009, article:Skillman_etal:2011}. Extended soft thermal X-ray (SXR) emission is evident from the ROSAT all-sky survey in the 0.1 to 2.4 keV band \citep{article:BrielHenryBohringer:1992} and observations with XMM-Newton \citep{article:Briel_etal:2001} revealed substructure in the X-ray halo with substantial turbulent pressure support that shows a lower limit of $\sim 10 \%$ of the total intracluster medium pressure \citep{article:Schuecker_etal:2004}. As such it is a natural candidate for gamma-ray observations.
